---
title: "phynotype"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

`phynotype` is an R package for clustering workflows, consensus meta-clustering, validation, exploration, prediction, and plotting.

`phynotype` is a clustering workflow package. It is not a phenotypic data processing package.

## Installation

`phynotype` is not on CRAN yet. Install the development version from GitHub:

```{r install, eval=FALSE}
# install.packages("pak")
pak::pkg_install("ielbadisy/phynotype")
```

Or install from a local source checkout:

```{r install-local, eval=FALSE}
pak::pkg_install("path/to/phynotype")
```

## Quick start

```{r quick-start}
library(phynotype)

fit <- cluster(iris[, 1:4], method = "kmeans", k = 3, seed = 1)
fit
summary(fit)
```

## Meta-clustering

```{r meta-clustering}
mfit <- metacluster(
  iris[, 1:4],
  methods = c("kmeans", "pam", "hclust"),
  k = 2:5,
  consensus = "coassoc",
  seed = 1
)

mfit
summary(mfit)
```

## Validation

```{r validation}
validate(fit)
validate(mfit)
validate(iris[, 1:4], method = "kmeans", k = 2:6, seed = 1)
```

## Exploration

```{r exploration}
exp <- explore(fit)
exp
head(exp$feature_summary)
```

## Plotting

```{r plot-clusters}
plot_clusters(fit)
```

```{r plot-silhouette}
plot_silhouette(fit)
```

```{r plot-consensus}
plot_consensus(mfit)
```

```{r plot-coassoc}
plot_coassoc(mfit)
```

```{r plot-feature-profiles}
plot_feature_profiles(explore(fit))
```

```{r plot-cluster-sizes}
plot_cluster_sizes(fit)
```

## Prediction

```{r prediction}
pred <- predict(fit, iris[1:10, 1:4])
pred
data.frame(
  observation = seq_along(pred$clusters),
  cluster = pred$clusters
)
```
