---
title: "phynotype"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

`phynotype` is an R package for clustering workflows, consensus meta-clustering, validation, exploration, prediction, and plotting.

`phynotype` is a clustering workflow package. It is not a phenotypic data processing package.

## Installation

`phynotype` is not on CRAN yet. Install the development version from GitHub:

```{r install, eval=FALSE}
# install.packages("pak")
pak::pkg_install("ielbadisy/phynotype")
```

Or install from a local source checkout:

```{r install-local, eval=FALSE}
pak::pkg_install("path/to/phynotype")
```

## Quick start

```{r quick-start}
library(phynotype)

fit <- cluster(iris[, 1:4], method = "kmeans", k = 3, seed = 1)
fit
summary(fit)
```

## API accessors

The main objects expose a common accessor API, so downstream code does not need
to inspect object internals.

```{r api-accessors}
method_used(fit)
n_clusters(fit)
data.frame(cluster = names(sizes(fit)), size = unname(sizes(fit)))
head(clusters(fit))
centers(fit)
```

## Meta-clustering

```{r meta-clustering}
mfit <- metacluster(
  iris[, 1:4],
  methods = c("kmeans", "pam", "hclust"),
  k = 2:5,
  consensus = "coassoc",
  seed = 1
)

mfit
summary(mfit)
```

Inspect the candidate solutions, consensus selection, and co-association matrix:

```{r meta-clustering-details}
head(mfit$candidate_table)
mfit$selection_summary
mfit$stability_summary
round(mfit$coassoc_matrix[1:6, 1:6], 2)
```

## Validation

```{r validation}
validate(fit)
validate(mfit)
validate(iris[, 1:4], method = "kmeans", k = 2:6, seed = 1)
```

## Exploration

```{r exploration}
exp <- explore(fit)
exp
head(exp$feature_summary)
```

## Plotting

```{r plot-clusters}
plot_clusters(fit)
```

```{r plot-silhouette}
plot_silhouette(fit)
```

```{r plot-consensus}
plot_consensus(mfit)
```

```{r plot-coassoc}
plot_coassoc(mfit)
```

```{r plot-feature-profiles}
plot_feature_profiles(explore(fit))
```

```{r plot-cluster-sizes}
plot_cluster_sizes(fit)
```

## Prediction

```{r prediction}
pred <- predict(fit, iris[1:10, 1:4])
pred
data.frame(
  observation = seq_along(pred$clusters),
  cluster = pred$clusters
)
```

## Interpretability

Global permutation importance estimates which features the fitted clustering
rule relies on most.

```{r feature-importance}
imp <- feature_importance(fit, n_repeats = 3, seed = 1)
imp
imp$summary
```

```{r plot-feature-importance}
plot(imp)
```

Ceteris paribus profiles show how local predictions change when one feature is
varied and the other features are held fixed.

```{r ceteris-paribus}
cp <- ceteris_paribus(
  fit,
  iris[1:2, 1:4],
  features = c("Petal.Length", "Petal.Width"),
  grid_size = 6,
  target = "score"
)

cp
head(cp$profiles)
```

```{r plot-ceteris-paribus}
plot(cp)
```

LIME-style explanations fit local surrogate models around selected
observations.

```{r lime-explain}
lx <- lime_explain(
  fit,
  iris[1:2, 1:4],
  n_features = 3,
  n_permutations = 50,
  seed = 1
)

lx
lx$explanations
```

```{r plot-lime-explain}
plot(lx)
```
