---
title: "phynotype API Demonstration"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
devtools::load_all(".")
set.seed(1)
iris_x <- iris[, 1:4]
iris_truth <- as.integer(iris$Species)
```

# Overview

This report demonstrates the main `phynotype` API:

- `cluster()`
- `metacluster()`
- `validate()`
- `explore()`
- `predict()`
- plotting helpers

# Single clustering

## K-means

```{r}
fit_kmeans <- cluster(
  iris_x,
  method = "kmeans",
  k = 3,
  scale = TRUE,
  seed = 1
)

fit_kmeans
summary(fit_kmeans)
clusters(fit_kmeans)[1:10]
sizes(fit_kmeans)
method_used(fit_kmeans)
```

## PAM

```{r}
fit_pam <- cluster(
  iris_x,
  method = "pam",
  k = 3,
  scale = TRUE,
  seed = 1
)

fit_pam
summary(fit_pam)
prototypes(fit_pam)
```

## Hierarchical clustering

```{r}
fit_hclust <- cluster(
  iris_x,
  method = "hclust",
  k = 3,
  scale = TRUE,
  seed = 1
)

fit_hclust
summary(fit_hclust)
```

## AGNES

```{r}
fit_agnes <- cluster(
  iris_x,
  method = "agnes",
  k = 3,
  scale = TRUE,
  seed = 1
)

fit_agnes
summary(fit_agnes)
```

## Gaussian mixture model

```{r}
if (requireNamespace("mclust", quietly = TRUE)) {
  fit_gmm <- cluster(
    iris_x,
    method = "gmm",
    k = 3,
    scale = TRUE,
    seed = 1
  )

  fit_gmm
  summary(fit_gmm)
  dim(membership(fit_gmm))
} else {
  "Package `mclust` is not installed, so the GMM example is skipped."
}
```

## DBSCAN

```{r}
if (requireNamespace("dbscan", quietly = TRUE)) {
  sim <- phynotype:::synthetic_clusters(n_per_cluster = 30, seed = 1)
  fit_dbscan <- cluster(
    sim$x,
    method = "dbscan",
    eps = 0.9,
    minPts = 5,
    scale = TRUE
  )

  fit_dbscan
  summary(fit_dbscan)
  table(clusters(fit_dbscan))
} else {
  "Package `dbscan` is not installed, so the DBSCAN example is skipped."
}
```

# Validation

## Validation for a fitted object

```{r}
val_kmeans <- validate(fit_kmeans, truth = iris_truth)
val_kmeans
val_kmeans$metrics_table
val_kmeans$per_cluster_table
```

## Validation over a grid of k

```{r}
val_grid <- validate(
  iris_x,
  method = "kmeans",
  k = 2:6,
  scale = TRUE,
  seed = 1
)

head(val_grid$metrics_table)
```

# Exploration

```{r}
exp_kmeans <- explore(fit_kmeans)
exp_kmeans
head(exp_kmeans$size_table)
head(exp_kmeans$feature_summary)
head(exp_kmeans$separation_table)
head(exp_kmeans$embedding)
```

# Prediction

## K-means prediction

```{r}
pred_kmeans <- predict(fit_kmeans, iris_x[1:10, ])
pred_kmeans
pred_kmeans$clusters
head(pred_kmeans$distances)
```

## PAM prediction

```{r}
pred_pam <- predict(fit_pam, iris_x[1:10, ])
pred_pam
pred_pam$clusters
```

## GMM prediction

```{r}
if (exists("fit_gmm")) {
  pred_gmm <- predict(fit_gmm, iris_x[1:10, ])
  pred_gmm
  pred_gmm$clusters
  head(pred_gmm$membership)
}
```

## Unsupported hierarchical prediction

```{r}
try(predict(fit_hclust, iris_x[1:5, ]))
```

# Meta-clustering

```{r}
mfit <- metacluster(
  iris_x,
  methods = c("kmeans", "pam", "hclust"),
  k = 2:5,
  consensus = "coassoc",
  scale = TRUE,
  seed = 1
)

mfit
summary(mfit)
head(mfit$candidate_table)
dim(mfit$coassoc_matrix)
head(mfit$selection_summary)
```

## Meta-cluster validation and exploration

```{r}
mval <- validate(mfit, truth = iris_truth)
mval$metrics_table

mexp <- explore(mfit)
head(mexp$feature_summary)
```

# Plots

## Cluster plots

```{r}
plot_clusters(fit_kmeans)
plot_cluster_sizes(fit_kmeans)
plot_feature_profiles(exp_kmeans)
plot_silhouette(fit_kmeans)
plot_validation(val_kmeans)
```

## Hierarchical plot

```{r}
plot_dendrogram(fit_hclust)
```

## Meta-cluster plots

```{r}
plot_consensus(mfit)
plot_coassoc(mfit)
plot_clusters(mfit)
plot_cluster_sizes(mfit)
plot_validation(mval)
plot_dendrogram(mfit)
```

# Session info

```{r}
sessionInfo()
```
