---
title: "Clustering with phynotype: A Complete Tutorial"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Clustering with phynotype: A Complete Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5,
  warning   = FALSE,
  message   = FALSE
)
set.seed(42)
```

# Introduction

**phynotype** is a unified R framework for unsupervised clustering workflows.
It provides a consistent grammar across seven clustering algorithms, robust
validation metrics, exploratory visualization, consensus meta-clustering, and
permutation feature importance — all through a handful of top-level functions:

| Function              | Purpose                                          |
|-----------------------|--------------------------------------------------|
| `cluster()`           | Fit a single clustering solution                 |
| `metacluster()`       | Fit a consensus meta-clustering solution         |
| `validate()`          | Compute internal / external validation metrics   |
| `explore()`           | Summarize sizes, profiles, and PCA embedding     |
| `predict()`           | Assign new observations to fitted clusters       |
| `feature_importance()`| Permutation feature importance                   |
| `prepare_mixed_data()`| Encode mixed-type data for numeric methods       |
| `mixed_distance()`    | Gower distance matrix for hierarchical methods   |

The tutorial works through each piece step by step, from mathematical
foundations to working code and result interpretation.

---

# Setup

```{r load}
library(phynotype)
library(ggplot2)

# Reproducible synthetic data --------------------------------------------------
sim_gaussian <- function(n = 150, seed = 42) {
  set.seed(seed)
  centers <- list(
    c(0,  0),
    c(5,  1),
    c(2,  6)
  )
  dat <- do.call(rbind, lapply(seq_along(centers), function(i) {
    x <- rnorm(n / length(centers), centers[[i]][1], 0.9)
    y <- rnorm(n / length(centers), centers[[i]][2], 0.9)
    data.frame(x = x, y = y, true_label = i)
  }))
  dat
}

raw      <- sim_gaussian()
X        <- as.matrix(raw[, c("x", "y")])
truth    <- raw$true_label
```

```{r plot-raw, fig.cap="Simulated data — three Gaussian clusters in 2D."}
ggplot(raw, aes(x, y, colour = factor(true_label))) +
  geom_point(size = 2) +
  labs(title = "Simulated data", colour = "True cluster") +
  theme_minimal()
```

---

# Supported Methods at a Glance

```{r method-table, echo=FALSE}
knitr::kable(
  data.frame(
    Method   = c("kmeans", "pam", "hclust", "agnes", "dbscan", "gmm", "kproto"),
    Type     = c("Partition", "Partition", "Hierarchical", "Hierarchical",
                 "Density", "Model-based", "Partition (mixed)"),
    Requires_k = c("Yes", "Yes", "Yes (cut)", "Yes (cut)", "No", "Yes", "Yes"),
    Mixed_data = c("No", "No", "No*", "No*", "No", "No", "Yes"),
    Key_param  = c("nstart, iter.max", "—", "linkage", "linkage",
                   "eps, minPts", "—", "lambda"),
    stringsAsFactors = FALSE
  ),
  caption = "* Use `mixed_distance()` to precompute a Gower distance first."
)
```

---

# K-Means Clustering

## Mathematical Formulation

K-means partitions $n$ observations into $k$ clusters
$\mathcal{C} = \{C_1, \ldots, C_k\}$ by minimising the total within-cluster
sum of squares (WCSS):

$$
\min_{\mathcal{C}} \sum_{j=1}^{k} \sum_{\mathbf{x}_i \in C_j}
\|\mathbf{x}_i - \boldsymbol{\mu}_j\|^2
$$

where $\boldsymbol{\mu}_j = \frac{1}{|C_j|}\sum_{\mathbf{x}_i \in C_j}\mathbf{x}_i$
is the centroid of cluster $j$.

**Algorithm (Lloyd's)**

1. Randomly initialise $k$ centroids $\{\boldsymbol{\mu}_1,\ldots,\boldsymbol{\mu}_k\}$.
2. **Assignment step.** Assign each observation to its nearest centroid:
$$c_i = \arg\min_{j} \|\mathbf{x}_i - \boldsymbol{\mu}_j\|^2$$
3. **Update step.** Recompute each centroid as the mean of its members.
4. Repeat steps 2–3 until assignments no longer change.

`cluster()` runs `nstart = 10` random initialisations by default and returns
the solution with the lowest WCSS.

## Fitting

```{r kmeans-fit}
fit_km <- cluster(X, method = "kmeans", k = 3, seed = 1)
fit_km
```

## Key Accessors

```{r kmeans-accessors}
clusters(fit_km)   |> table()   # cluster labels
centers(fit_km)                 # centroid matrix
n_clusters(fit_km)              # number of clusters
sizes(fit_km)                   # per-cluster size
method_used(fit_km)             # "kmeans"
```

## Elbow Plot (choosing k)

A common heuristic is to plot WCSS against $k$ and look for the "elbow":

```{r kmeans-elbow, fig.cap="Elbow plot: WCSS vs. number of clusters."}
wcss <- vapply(2:7, function(k) {
  fit <- cluster(X, method = "kmeans", k = k, seed = 1)
  fit$extras$tot_withinss
}, numeric(1))

ggplot(data.frame(k = 2:7, wcss = wcss), aes(k, wcss)) +
  geom_line() + geom_point(size = 3) +
  scale_x_continuous(breaks = 2:7) +
  labs(title = "Elbow plot", y = "Total within-cluster SS", x = "k") +
  theme_minimal()
```

## Visualisation

```{r kmeans-plot, fig.cap="K-means solution projected onto PC1 / PC2."}
plot_clusters(fit_km)
```

---

# PAM — Partitioning Around Medoids

## Mathematical Formulation

PAM minimises the total sum of **dissimilarities** to the nearest medoid:

$$
\min_{m_1,\ldots,m_k} \sum_{i=1}^{n} \min_{j} d(\mathbf{x}_i, m_j)
$$

where $m_j$ is an actual observation (medoid) of cluster $j$ and
$d(\cdot,\cdot)$ is a user-supplied dissimilarity (default: Euclidean).

Unlike k-means, PAM is:

- **Robust to outliers** (medoids are real data points).
- Applicable to **any dissimilarity measure**.

**BUILD phase** — iteratively swap each non-medoid observation in if it
reduces total cost.
**SWAP phase** — iteratively swap medoids with non-medoids while total cost
decreases.

## Fitting

```{r pam-fit}
fit_pam <- cluster(X, method = "pam", k = 3, seed = 1)
fit_pam
```

```{r pam-prototypes}
prototypes(fit_pam)    # actual medoid observations
```

## Visualisation

```{r pam-plot, fig.cap="PAM solution in PCA space."}
plot_clusters(fit_pam)
```

---

# Hierarchical Clustering — `hclust` and `agnes`

## Mathematical Formulation

Hierarchical clustering builds a dendrogram by iteratively merging the two
most similar clusters. The linkage function determines how inter-cluster
distance is defined:

| Linkage      | Formula                                                              |
|--------------|----------------------------------------------------------------------|
| Single       | $d(A,B) = \min_{a \in A, b \in B} d(a,b)$                          |
| Complete     | $d(A,B) = \max_{a \in A, b \in B} d(a,b)$                          |
| Average (UPGMA) | $d(A,B) = \frac{1}{|A||B|}\sum_{a \in A}\sum_{b \in B}d(a,b)$   |
| Ward.D2      | $d(A,B) = \frac{|A||B|}{|A|+|B|}\|\bar{A}-\bar{B}\|^2$            |

The tree is cut at height $k$ to produce a flat partition of $k$ clusters.

**hclust** uses R's base `stats::hclust()`; **agnes** uses `cluster::agnes()`
which also provides the **agglomerative coefficient** (AC) — a measure of
clustering structure strength (AC near 1 means a clear hierarchy).

## Fitting

```{r hclust-fit}
# Build a distance matrix first (recommended practice)
D <- dist(X)

fit_hc <- cluster(D, method = "hclust", k = 3, linkage = "ward.D2")
fit_hc
```

```{r agnes-fit}
fit_ag <- cluster(X, method = "agnes", k = 3, linkage = "average")
fit_ag
```

## Dendrogram

```{r hclust-dendrogram, fig.cap="Ward.D2 dendrogram with cut at k=3."}
plot_dendrogram(fit_hc)
abline(h = 3.5, col = "red", lty = 2)   # illustrative cut height
```

## Linkage Comparison

```{r linkage-compare, fig.cap="PCA projections for four linkage strategies."}
linkages <- c("single", "complete", "average", "ward.D2")
plots <- lapply(linkages, function(lnk) {
  fit <- cluster(X, method = "hclust", k = 3, linkage = lnk)
  plot_clusters(fit) +
    labs(title = paste("hclust —", lnk)) +
    theme(legend.position = "none")
})

for (plt in plots) print(plt)
```

---

# DBSCAN — Density-Based Spatial Clustering

## Mathematical Formulation

DBSCAN groups observations that are **densely reachable** from each other and
marks the rest as **noise** (cluster label 0).

**Key definitions** (given $\varepsilon > 0$ and $\text{minPts} \ge 1$):

- **$\varepsilon$-neighbourhood** of $p$:
$N_\varepsilon(p) = \{q \mid d(p,q) \le \varepsilon\}$

- **Core point**: $|N_\varepsilon(p)| \ge \text{minPts}$

- **Directly density-reachable**: $q \in N_\varepsilon(p)$ and $p$ is a core
point.

- **Density-reachable**: there exists a chain
$p = p_1, p_2, \ldots, p_n = q$ where each $p_{i+1}$ is directly
density-reachable from $p_i$.

- **Density-connected**: both $p$ and $q$ are density-reachable from a common
point $o$.

A **cluster** is a maximal set of density-connected points.

**Choosing parameters:**

1. Plot the **k-NN distance** (sorted in ascending order). The "elbow" of
this curve is a good estimate for $\varepsilon$.
2. Set `minPts` $\ge$ the number of dimensions plus one (a common default is
`minPts = 5`).

## k-NN Distance Plot

```{r dbscan-knn, fig.cap="Sorted 5-NN distances — elbow near epsilon = 0.8."}
knn_dist <- sort(dbscan::kNNdist(X, k = 5))
ggplot(data.frame(idx = seq_along(knn_dist), d = knn_dist),
       aes(idx, d)) +
  geom_line() +
  geom_hline(yintercept = 0.8, colour = "red", linetype = 2) +
  labs(title = "5-NN distances (sorted)", x = "Observation rank",
       y = "Distance to 5th nearest neighbour") +
  theme_minimal()
```

## Fitting

```{r dbscan-fit}
fit_db <- cluster(X, method = "dbscan", eps = 0.8, minPts = 5)
fit_db
```

```{r dbscan-noise}
table(clusters(fit_db))   # 0 = noise points
```

```{r dbscan-plot, fig.cap="DBSCAN clusters (noise points appear as cluster 0 if any)."}
plot_clusters(fit_db)
```

---

# Gaussian Mixture Models (GMM)

## Mathematical Formulation

GMM assumes the data are generated by a mixture of $k$ Gaussian distributions:

$$
p(\mathbf{x}) = \sum_{j=1}^{k} \pi_j \,
\mathcal{N}(\mathbf{x} \mid \boldsymbol{\mu}_j, \boldsymbol{\Sigma}_j)
$$

where $\pi_j > 0$, $\sum_j \pi_j = 1$, $\boldsymbol{\mu}_j$ is the mean, and
$\boldsymbol{\Sigma}_j$ is the covariance of component $j$.

Parameters are estimated by **Expectation-Maximisation (EM)**:

**E-step** — compute posterior responsibilities:
$$
r_{ij} = \frac{\pi_j \,\mathcal{N}(\mathbf{x}_i\mid\boldsymbol{\mu}_j,\boldsymbol{\Sigma}_j)}
{\sum_{l=1}^{k}\pi_l\,\mathcal{N}(\mathbf{x}_i\mid\boldsymbol{\mu}_l,\boldsymbol{\Sigma}_l)}
$$

**M-step** — update parameters:
$$
\pi_j = \frac{1}{n}\sum_{i=1}^{n}r_{ij}, \quad
\boldsymbol{\mu}_j = \frac{\sum_i r_{ij}\mathbf{x}_i}{\sum_i r_{ij}}, \quad
\boldsymbol{\Sigma}_j = \frac{\sum_i r_{ij}(\mathbf{x}_i-\boldsymbol{\mu}_j)(\mathbf{x}_i-\boldsymbol{\mu}_j)^\top}{\sum_i r_{ij}}
$$

Model selection (covariance structure) is done via BIC.

**Soft assignments**: each observation gets a full probability vector
(accessible via `membership()`).

## Fitting

```{r gmm-fit}
fit_gmm <- cluster(X, method = "gmm", k = 3, seed = 1)
fit_gmm
```

```{r gmm-membership}
head(round(membership(fit_gmm), 3))   # posterior responsibilities
```

```{r gmm-plot, fig.cap="GMM clusters in PCA space."}
plot_clusters(fit_gmm)
```

---

# k-Prototypes — Mixed-Type Data

## Mathematical Formulation

k-prototypes extends k-means to handle both numeric and categorical features.
The dissimilarity between observation $\mathbf{x}$ and prototype $\mathbf{p}$
is:

$$
d(\mathbf{x}, \mathbf{p}) =
\underbrace{\sum_{r \in \text{num}} (x_r - p_r)^2}_{\text{numeric (Euclidean)}}
+
\underbrace{\lambda \sum_{c \in \text{cat}} \mathbf{1}[x_c \neq p_c]}_{\text{categorical (simple matching)}}
$$

The weight $\lambda > 0$ balances the two components. By default phynotype
lets `clustMixType::kproto()` choose $\lambda$ automatically.

A **prototype** for each cluster is a pair $(\boldsymbol{\mu}_j, \mathbf{m}_j)$
where $\boldsymbol{\mu}_j$ is the mean of numeric features and $\mathbf{m}_j$
is the mode vector of categorical features.

## Mixed Dataset

```{r kproto-data}
set.seed(1)
mix_df <- data.frame(
  age     = rnorm(120, mean = 45, sd = 12),
  bmi     = rnorm(120, mean = 26, sd = 4),
  smoking = sample(c("never", "former", "current"), 120, replace = TRUE),
  sex     = sample(c("M", "F"), 120, replace = TRUE)
)
head(mix_df)
```

## Fitting

```{r kproto-fit}
fit_kp <- cluster(mix_df, method = "kproto", k = 3, seed = 1)
fit_kp
```

```{r kproto-prototypes}
prototypes(fit_kp)   # mixed-type prototype per cluster
```

---

# Validation

## Internal Metrics — Mathematical Definitions

### Silhouette Width

For observation $i$ in cluster $C_j$, let:

- $a(i) = \frac{1}{|C_j|-1}\sum_{l \in C_j, l\neq i} d(i,l)$ — mean
intra-cluster distance
- $b(i) = \min_{j' \neq j} \frac{1}{|C_{j'}|}\sum_{l \in C_{j'}} d(i,l)$ —
mean distance to nearest cluster

$$
s(i) = \frac{b(i) - a(i)}{\max\{a(i),\, b(i)\}} \in [-1, 1]
$$

The **mean silhouette width** $\bar{s}$ is the average over all observations.
Values close to 1 indicate dense, well-separated clusters; values near 0
indicate overlapping clusters; negative values suggest misclassification.

### Calinski–Harabász Index

$$
\text{CH}(k) =
\frac{\text{BSS}/(k-1)}{\text{WSS}/(n-k)},
\quad
\text{BSS} = \sum_{j=1}^{k}|C_j|\|\boldsymbol{\mu}_j - \bar{\mathbf{x}}\|^2,
\quad
\text{WSS} = \sum_{j=1}^{k}\sum_{i \in C_j}\|\mathbf{x}_i - \boldsymbol{\mu}_j\|^2
$$

**Higher is better.** CH measures the ratio of between-cluster dispersion to
within-cluster dispersion.

### Davies–Bouldin Index

$$
\text{DB} = \frac{1}{k}\sum_{j=1}^{k} \max_{j' \neq j}
\frac{s_j + s_{j'}}{d(\boldsymbol{\mu}_j,\,\boldsymbol{\mu}_{j'})}
$$

where $s_j = \frac{1}{|C_j|}\sum_{i \in C_j}\|\mathbf{x}_i-\boldsymbol{\mu}_j\|$
is the average scatter of cluster $j$.

**Lower is better.** DB measures the average ratio of cluster scatter to
inter-cluster separation.

### Total Within-Cluster Sum of Squares

$$
\text{WCSS} = \sum_{j=1}^{k}\sum_{i \in C_j}\|\mathbf{x}_i - \boldsymbol{\mu}_j\|^2
$$

**Lower is better.** The objective function of k-means.

## External Metrics

External metrics require ground-truth labels $\mathbf{y}$.

### Adjusted Rand Index (ARI)

The Rand Index counts the fraction of observation pairs that are consistently
grouped (both in the same cluster, or both in different clusters). ARI adjusts
for chance:

$$
\text{ARI} = \frac{\sum_{ij}\binom{n_{ij}}{2} - \left[\sum_i\binom{a_i}{2}\sum_j\binom{b_j}{2}\right]/\binom{n}{2}}
{\frac{1}{2}\left[\sum_i\binom{a_i}{2}+\sum_j\binom{b_j}{2}\right]
- \left[\sum_i\binom{a_i}{2}\sum_j\binom{b_j}{2}\right]/\binom{n}{2}}
$$

$\text{ARI} = 1$ is a perfect match; $\text{ARI} \approx 0$ is random.

### Normalised Mutual Information (NMI)

$$
\text{NMI}(\mathbf{c},\mathbf{y}) =
\frac{I(\mathbf{c};\mathbf{y})}{\sqrt{H(\mathbf{c})H(\mathbf{y})}}
$$

where $I$ is mutual information and $H$ is Shannon entropy.
NMI $\in [0,1]$; 1 is perfect agreement.

## Bootstrap Stability (ARI-based)

For methods with a fixed $k$, phynotype estimates stability by:

1. Drawing $B$ bootstrap samples $X^*_b$ from $X$.
2. Refitting the model on each $X^*_b$.
3. Computing ARI between the reference partition and the bootstrap partition
on the overlapping observations.

The reported `bootstrap_ari` is the mean across $B$ bootstrap replicates.
Values near 1 indicate a stable solution that would be reproduced on new draws
from the same distribution.

The validation table also records the scale and preferred direction for each
metric so the output remains interpretable when multiple criteria are reported
together.

## Using `validate()`

```{r validate-single}
val_km <- validate(fit_km, truth = truth, n_boot = 20)
val_km
```

```{r validate-plot, fig.cap="Metric-specific validation panels for the k-means fit."}
plot_validation(val_km)
```

### Silhouette Plot

```{r silhouette-plot, fig.cap="Per-observation silhouette widths coloured by cluster."}
plot_silhouette(fit_km)
```

### Grid Validation — Choosing k Automatically

`validate()` applied directly to data runs the algorithm over a range of $k$
values and returns all metrics in a single table:

```{r validate-grid}
val_grid <- validate(X, method = "kmeans", k = 2:6, n_boot = 5)
val_grid$metrics_table
```

```{r validate-grid-plot, fig.cap="Validation metrics across k=2:6 for k-means."}
library(dplyr)
val_grid$metrics_table |>
  filter(metric %in% c("silhouette", "calinski_harabasz", "davies_bouldin")) |>
  ggplot(aes(k, value, colour = metric, group = metric)) +
  geom_line() + geom_point() +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "Validation metrics vs. k", x = "Number of clusters k") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

# Exploration — `explore()`

`explore()` returns a structured summary with four components:

| Component         | Content                                              |
|-------------------|------------------------------------------------------|
| `size_table`      | Observation count per cluster                        |
| `feature_summary` | Per-cluster feature means (and SD)                   |
| `separation_table`| Pairwise inter-cluster distances                     |
| `embedding`       | PCA coordinates for 2D plotting                      |

```{r explore}
exp_km <- explore(fit_km)
exp_km$size_table
exp_km$feature_summary
exp_km$separation_table
```

```{r plot-sizes, fig.cap="Cluster size bar chart."}
plot_cluster_sizes(fit_km)
```

```{r plot-profiles, fig.cap="Mean feature value per cluster (feature profiles)."}
plot_feature_profiles(exp_km)
```

---

# Prediction — `predict()`

After fitting on training data you can assign **new observations** to the
nearest centroid (k-means, hclust, agnes), nearest medoid (PAM), nearest
prototype (k-prototypes), or use the trained density model (GMM, DBSCAN).

## Distance-to-Centroid Assignment (k-means, PAM, hclust)

For k-means, new observation $\mathbf{x}^*$ is assigned to:

$$
\hat{c}(\mathbf{x}^*) = \arg\min_{j} \|\mathbf{x}^* - \boldsymbol{\mu}_j\|
$$

```{r predict}
new_obs <- matrix(c(1.0, 0.5,   # near cluster 1
                    5.2, 1.1,   # near cluster 2
                    2.1, 6.3),  # near cluster 3
                  nrow = 3, byrow = TRUE,
                  dimnames = list(NULL, c("x", "y")))

pred <- predict(fit_km, new_data = new_obs)
pred
pred$clusters
pred$distances   # distance to each centroid
```

---

# Meta-Clustering — `metacluster()`

## Motivation

A single clustering run is sensitive to the choice of algorithm, $k$, and
random initialisation. **Meta-clustering** aggregates multiple candidate
partitions into a single, robust consensus solution.

## Co-Association Matrix

Given $M$ candidate partitions $\{\mathbf{c}^{(1)}, \ldots, \mathbf{c}^{(M)}\}$,
the **co-association matrix** $C \in [0,1]^{n \times n}$ records the
proportion of partitions that place observations $i$ and $j$ in the same
cluster:

$$
C_{ij} = \frac{1}{M}\sum_{m=1}^{M}\mathbf{1}\{c_i^{(m)} = c_j^{(m)}\}
$$

$C_{ij}$ close to 1 means the two observations are consistently co-clustered;
close to 0 means they are consistently separated. This gives a
**consensus dissimilarity**:

$$
D = 1 - C
$$

## Consensus Partition

$D$ is passed to **hierarchical clustering** (average linkage) to produce a
dendrogram. The optimal cut $k^*$ is chosen by maximising the mean silhouette
width on $D$:

$$
k^* = \arg\max_{k} \bar{s}(k, D)
$$

## Stability — Pairwise Partition Agreement

Candidate-level stability is measured by the fraction of observation **pairs**
that are treated consistently across any two candidate partitions
$(\mathbf{c}^{(m)},\mathbf{c}^{(m')})$:

$$
\text{PPA}(m,m') =
\frac{1}{n^2}\sum_{i=1}^{n}\sum_{j=1}^{n}
\mathbf{1}\{[c_i^{(m)} = c_j^{(m)}] = [c_i^{(m')} = c_j^{(m')}]\}
$$

The reported `mean_agreement` is the average PPA across all $\binom{M}{2}$
pairs.

## Fitting

```{r metacluster-fit}
mfit <- metacluster(
  X,
  methods = c("kmeans", "pam", "hclust"),
  k       = 2:4,
  seed    = 1
)
mfit
```

```{r metacluster-k}
n_clusters(mfit)         # consensus k selected
mfit$selection_summary   # silhouette scores per candidate k
mfit$stability_summary   # pairwise partition agreement
```

## Visualisations

```{r coassoc-plot, fig.cap="Co-association matrix — dark squares mark consistently co-clustered pairs."}
plot_coassoc(mfit)
```

```{r consensus-dendrogram, fig.cap="Consensus dendrogram built from the dissimilarity D = 1 - C."}
plot_dendrogram(mfit)
```

```{r consensus-clusters, fig.cap="Consensus cluster assignments in PCA space."}
plot_clusters(mfit)
```

## Validation of a Meta-Clustering Solution

```{r metacluster-validate}
val_meta <- validate(mfit, truth = truth)
val_meta
```

---

# Mixed-Data Preprocessing

## `prepare_mixed_data()`

Encodes a mixed-type data frame into a numeric matrix for use with k-means,
PAM, GMM, or DBSCAN.

- **Numeric columns**: optionally centred and scaled.
- **Categorical / logical / character columns**: one-hot encoded (full-rank by
default).
- **Missing values**: numeric columns are median-imputed; categorical columns
gain an explicit `"missing"` level.

```{r prepare-mixed}
mat <- prepare_mixed_data(
  mix_df,
  center     = TRUE,
  scale      = TRUE,
  drop_first = FALSE,
  na_action  = "keep"
)
dim(mat)
head(mat[, 1:6])
```

```{r prepare-mixed-attr}
attr(mat, "phynotype_preprocess")$encoded_map
```

Now `mat` can be passed directly to `cluster()` with any numeric method:

```{r prepare-mixed-cluster}
fit_mixed_km <- cluster(mat, method = "kmeans", k = 3, seed = 1)
sizes(fit_mixed_km)
```

## `mixed_distance()`

For hierarchical methods on mixed data, compute a **Gower distance** matrix:

$$
d_\text{Gower}(\mathbf{x}_i, \mathbf{x}_j) =
\frac{\sum_{r=1}^{p} w_r \, \delta_r(x_{ir}, x_{jr}) \, d_r(x_{ir}, x_{jr})}
{\sum_{r=1}^{p} w_r \, \delta_r(x_{ir}, x_{jr})}
$$

where $\delta_r = 0$ if either value is missing, $d_r = |x_{ir}-x_{jr}|/R_r$
for numeric variables (range-normalised), and $d_r = \mathbf{1}[x_{ir}\neq x_{jr}]$
for categorical variables.

```{r mixed-distance}
D_gower <- mixed_distance(mix_df, metric = "gower")
fit_hc_mix <- cluster(D_gower, method = "hclust", k = 3, linkage = "average")
sizes(fit_hc_mix)
```

---

# Feature Importance

## Mathematical Formulation

Permutation importance breaks the marginal information in feature $j$ by
randomly shuffling its values, then measures how much the fitted clustering
rule changes. For the default **instability** metric:

$$
\text{FI}_j = \frac{1}{M}\sum_{m=1}^{M}
\frac{1}{n}\sum_{i=1}^{n}
\mathbf{1}\{\hat{c}_i \neq f(X^{(m)}_{\pi(j)})_i\}
$$

where $X^{(m)}_{\pi(j)}$ denotes the data with feature $j$ permuted in
repetition $m$, and $f(\cdot)$ is the fitted prediction rule.

For the **silhouette** metric:

$$
\text{FI}_j = \frac{1}{M}\sum_{m=1}^{M}
\left[\bar{s}(X) - \bar{s}(X^{(m)}_{\pi(j)})\right]
$$

For **total within-cluster SS**:

$$
\text{FI}_j = \frac{1}{M}\sum_{m=1}^{M}
\left[W(X^{(m)}_{\pi(j)}) - W(X)\right]
$$

Larger values always indicate greater importance.

## Using `feature_importance()`

```{r fi-instability, fig.cap="Permutation feature importance (instability metric)."}
imp <- feature_importance(fit_km, metric = "instability", n_repeats = 20, seed = 1)
imp
plot(imp)
```

```{r fi-silhouette, fig.cap="Feature importance measured by loss in mean silhouette."}
imp_sil <- feature_importance(fit_km, metric = "silhouette", n_repeats = 20, seed = 1)
plot(imp_sil)
```

```{r fi-within, fig.cap="Feature importance measured by increase in WCSS."}
imp_wss <- feature_importance(fit_km, metric = "total_within", n_repeats = 20, seed = 1)
plot(imp_wss)
```

---

# End-to-End Workflow Summary

```r
library(phynotype)

# 1. Load / simulate data
X     <- as.matrix(iris[, 1:4])
truth <- as.integer(iris$Species)

# 2. Preprocess (mixed data only — skip for numeric)
# mat <- prepare_mixed_data(df, center = TRUE, scale = TRUE)

# 3. Fit a single method
fit <- cluster(X, method = "kmeans", k = 3, seed = 1)

# 4. Validate (internal + external + stability)
val <- validate(fit, truth = truth, n_boot = 20)
print(val)
plot_validation(val)
plot_silhouette(fit)

# 5. Explore structure
exp <- explore(fit)
plot_clusters(fit)
plot_cluster_sizes(fit)
plot_feature_profiles(exp)

# 6. Predict new observations
pred <- predict(fit, new_data = X[1:5, ])

# 7. Assess feature importance
imp <- feature_importance(fit, metric = "instability", n_repeats = 10, seed = 1)
plot(imp)

# 8. Consensus meta-clustering
mfit <- metacluster(X, methods = c("kmeans","pam","hclust"), k = 2:5, seed = 1)
plot_coassoc(mfit)
plot_dendrogram(mfit)
plot_clusters(mfit)
validate(mfit, truth = truth)
```

---

# Session Info

```{r session}
sessionInfo()
```
