---
title: "Interpreting clustering solutions with phynotype"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting clustering solutions with phynotype}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(phynotype)
```

## Why clustering explanations need a separate workflow

Supervised model explanations usually target a known outcome. Clustering does
not have an observed response. The object being explained is the fitted
partitioning rule: how a clustering algorithm maps a row of features to a
cluster label, a membership probability, or a prototype similarity score.

This distinction matters. A clustering explanation does not say that a feature
causes a clinical phenotype or biological subgroup. It says that the fitted
clustering solution changes in a specific way when the feature is perturbed,
held fixed, or locally approximated. The interpretation is therefore conditional
on the preprocessing, algorithm, distance geometry, and selected number of
clusters.

`phynotype` provides three complementary interpretability methods:

1. `feature_importance()` for global permutation importance.
2. `ceteris_paribus()` for individual conditional profiles.
3. `lime_explain()` for sparse local surrogate explanations.

The methods share a model-agnostic design. They call `predict()` on a
`cluster_fit` object, so they apply to clustering engines with native prediction
support, such as k-means, PAM, Gaussian mixture models, and DBSCAN when the
installed backend supports prediction. Hierarchical fits created from
row-by-feature data are interpreted through nearest-centroid prediction from the
cut tree. Mixed-data k-prototypes fits are supported for permutation
importance, ceteris paribus profiles, and local surrogate explanations. Fits
created only from a precomputed distance object cannot be interpreted for new
rows because the original feature space is not available.

## Fit a clustering model

The examples use the numeric iris measurements as a compact demonstration data
set.

```{r fit}
x <- iris[, 1:4]
fit <- cluster(x, method = "kmeans", k = 3, scale = TRUE, seed = 7)
fit
```

## Global permutation feature importance

Permutation importance asks how much the fitted clustering behavior degrades
when the information in one feature is broken. For feature \(j\), repetition
\(m\), and permuted data matrix \(X^{(m)}_{\pi(j)}\), the default instability
importance is

\[
FI_j =
\frac{1}{M}\sum_{m=1}^{M}
\frac{1}{n}\sum_{i=1}^{n}
I\left\{\hat{c}_i \neq f(X^{(m)}_{\pi(j)})_i\right\},
\]

where \(\hat{c}=f(X)\) is the baseline predicted partition. A value near zero
means that permuting the feature rarely changes assignments. A larger value
means that the feature is influential for the fitted assignment rule.

```{r feature-importance}
imp <- feature_importance(
  fit,
  n_repeats = 5,
  seed = 1
)

imp
imp$summary
```

```{r plot-feature-importance}
plot(imp)
```

The same interface can use internal clustering criteria. For silhouette, larger
is better, so importance is the loss of score:

\[
FI_j =
\frac{1}{M}\sum_{m=1}^{M}
\left[
S(f, X) - S(f, X^{(m)}_{\pi(j)})
\right].
\]

For total within-cluster sum of squares, smaller is better, so importance is the
increase in dispersion:

\[
FI_j =
\frac{1}{M}\sum_{m=1}^{M}
\left[
W(f, X^{(m)}_{\pi(j)}) - W(f, X)
\right].
\]

```{r feature-importance-within}
feature_importance(
  fit,
  metric = "total_within",
  n_repeats = 3,
  seed = 1
)$summary
```

Custom losses are useful when a project has a domain-specific objective. A
custom loss must return a scalar where larger values are worse.

```{r custom-loss}
balance_loss <- function(data, clusters, object) {
  tab <- table(clusters)
  stats::sd(as.numeric(tab))
}

feature_importance(
  fit,
  loss = balance_loss,
  n_repeats = 3,
  seed = 1
)$summary
```

## Ceteris paribus profiles

Ceteris paribus profiles vary one feature while holding all others fixed. For
observation \(x_i\), feature \(j\), and grid value \(z\), the profile is

\[
CP_{i,j}(z) = g\{f(x_{i,-j}, z)\},
\]

where \(g\) extracts either the predicted cluster label or a cluster-specific
score. This is an individual conditional diagnostic: it shows how the fitted
rule behaves around a concrete profile.

```{r cp-cluster}
cp <- ceteris_paribus(
  fit,
  new_data = x[1:3, ],
  features = c("Sepal.Length", "Petal.Length"),
  grid_size = 15,
  target = "cluster"
)

head(cp$profiles)
```

```{r plot-cp-cluster}
plot(cp)
```

For distance-based methods such as k-means and PAM, `phynotype` converts
distances to prototypes into normalized radial similarity scores. These scores
are not calibrated probabilities; they are smooth local summaries that make
profiles easier to compare.

```{r cp-score}
cp_score <- ceteris_paribus(
  fit,
  new_data = x[1:2, ],
  features = "Petal.Width",
  grid_size = 20,
  target = "score"
)

head(cp_score$profiles)
```

## LIME-style local surrogate explanations

LIME explains one observation by fitting a simple surrogate model in a local
neighborhood. Let \(z_i\) be perturbed samples around the observation \(x\),
\(\pi_x(z_i)\) proximity weights, \(g\{f(z_i)\}\) the clustering target, and
\(h\) a sparse local surrogate. The objective is

\[
\hat{h}
=
\arg\min_{h \in H}
\sum_{i=1}^{N}
\pi_x(z_i)
\left[g\{f(z_i)\} - h(z_i)\right]^2
+ \Omega(h).
\]

`phynotype` uses a weighted linear surrogate and returns the largest local
standardized coefficients. For `target = "cluster"`, the response is a local
one-vs-cluster assignment indicator. For `target = "score"`, the response is a
cluster-specific membership or similarity score.

```{r lime}
lx <- lime_explain(
  fit,
  new_data = x[1:3, ],
  n_features = 3,
  n_permutations = 100,
  seed = 2
)

lx
lx$explanations
```

```{r plot-lime}
plot(lx)
```

The signs of the local coefficients describe the surrogate, not the original
algorithm directly. A positive coefficient means that, within the generated
local neighborhood and after standardization, larger values of the feature are
associated with a higher local target value.

## Case study: interpreting an iris clustering solution

This case study combines the three diagnostics into one reporting workflow. The
goal is not to recover the species labels as a supervised target. Species are
used only after fitting to help audit whether the unsupervised partition has a
recognizable structure.

```{r case-study-fit}
case_x <- iris[, 1:4]
case_fit <- cluster(case_x, method = "kmeans", k = 3, scale = TRUE, seed = 11)

table(cluster = clusters(case_fit), species = iris$Species)
validate(case_fit, truth = iris$Species, n_boot = 3)$metrics_table
```

Global permutation importance identifies the measurements that the fitted rule
uses most strongly.

```{r case-study-importance}
case_importance <- feature_importance(
  case_fit,
  n_repeats = 10,
  seed = 11
)

case_importance$summary
plot(case_importance)
```

The strongest features can then be inspected locally. The following profiles
explain one representative row from each fitted cluster and show how the
cluster score changes when petal measurements are varied one at a time.

```{r case-study-cp}
representatives <- tapply(seq_len(nrow(case_x)), clusters(case_fit), function(idx) idx[1])
case_profiles <- ceteris_paribus(
  case_fit,
  new_data = case_x[unlist(representatives), ],
  features = c("Petal.Length", "Petal.Width"),
  grid_size = 25,
  target = "score"
)

plot(case_profiles)
```

Finally, local surrogate explanations are useful for concrete profiles. Here
the first row from each fitted cluster is explained for its predicted cluster.

```{r case-study-lime}
case_lime <- lime_explain(
  case_fit,
  new_data = case_x[unlist(representatives), ],
  n_features = 3,
  n_permutations = 200,
  seed = 12
)

case_lime$explanations
plot(case_lime)
```

Taken together, the case study gives three complementary checks: the global
importance ranking shows which features drive the partition, ceteris paribus
profiles show whether cluster scores change smoothly around representative
observations, and LIME-style surrogates summarize the local direction of
effects for individual rows.

## Parallel execution with functionals

Permutation repeats, ceteris paribus profiles, and LIME neighborhoods are
independent tasks. When the suggested `functionals` package is installed, the
interpretability functions can dispatch those tasks through `functionals::fmap()`
with optional progress reporting.

```{r parallel, eval = FALSE}
imp_parallel <- feature_importance(
  fit,
  n_repeats = 25,
  seed = 1,
  parallel = TRUE,
  workers = 2,
  progress = TRUE
)

cp_parallel <- ceteris_paribus(
  fit,
  new_data = x[1:10, ],
  features = colnames(x),
  parallel = TRUE,
  workers = 2
)

lx_parallel <- lime_explain(
  fit,
  new_data = x[1:10, ],
  n_permutations = 500,
  parallel = TRUE,
  workers = 2
)
```

If `functionals` is unavailable, `phynotype` falls back to sequential execution
with a warning. This keeps the package usable in standard CRAN environments
while still supporting parallel computation in local analysis workflows.

## Recommended reporting workflow

A defensible clustering interpretation should combine global and local checks.
One practical sequence is:

1. Fit the clustering solution and report preprocessing, method, number of
   clusters, and validation metrics.
2. Use permutation feature importance to identify globally influential features.
3. Use ceteris paribus profiles to inspect whether influential features change
   assignments smoothly, abruptly, or only in narrow ranges.
4. Use LIME explanations on representative observations, borderline cases, and
   clinically or scientifically important profiles.
5. Report instability due to stochastic algorithms by repeating the analysis
   with fixed seeds and, when appropriate, alternative clustering methods.

## Limitations

These methods explain the fitted clustering rule, not an external truth. They
can be affected by correlated features, scaling choices, cluster imbalance, and
regions of the feature space with little observed data. In particular,
permutation can create unrealistic feature combinations, ceteris paribus
profiles can move an observation outside the empirical manifold, and LIME
surrogates depend on the perturbation distribution and kernel width.

Interpretability results should therefore be treated as diagnostics for the
stability and geometry of a fitted solution. They are strongest when reported
alongside validation metrics, sensitivity analyses, and substantive domain
knowledge.
