---
title: "phynotype: complete workflow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{phynotype: complete workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(phynotype)
```

`phynotype` provides one workflow for clustering, validation, exploration,
prediction, and interpretation. This vignette is the canonical package
walkthrough and replaces the older split articles.

## Algorithms

### K-means

Minimize
\[
\min_{\mathcal C} \sum_{j=1}^k \sum_{x_i \in C_j} \lVert x_i - \mu_j \rVert^2.
\]

```{r}
fit_kmeans <- cluster(iris[, 1:4], method = "kmeans", k = 3, seed = 1)
fit_kmeans
```

### PAM

Minimize the dissimilarity to medoids:
\[
\min_{m_1,\ldots,m_k} \sum_{i=1}^n \min_j d(x_i, m_j).
\]

```{r}
fit_pam <- cluster(iris[, 1:4], method = "pam", k = 3, seed = 1)
fit_pam
```

### Hierarchical clustering

`hclust` and `agnes` work on numeric matrices or on `dist` objects created by
`mixed_distance()`.

```{r}
d <- mixed_distance(iris[, 1:4])
fit_hclust <- cluster(d, method = "hclust", k = 3)
fit_hclust
```

### DBSCAN

```{r}
fit_dbscan <- cluster(iris[, 1:4], method = "dbscan", eps = 0.5, minPts = 5)
fit_dbscan
```

### Gaussian mixture models

```{r}
fit_gmm <- cluster(iris[, 1:4], method = "gmm", k = 3, seed = 1)
fit_gmm
```

### K-prototypes

For mixed data, `kproto` combines numeric centers and categorical modes.

```{r}
fit_kproto <- cluster(iris, method = "kproto", k = 3, seed = 1)
fit_kproto
```

### KMM

KMM minimizes a weighted mixed-distance criterion:
\[
D(x_i, p_g) = \sum_{j \in \mathcal N} (x_{ij} - \mu_{gj})^2 +
\lambda \sum_{j \in \mathcal C} I(x_{ij} \neq \nu_{gj}),
\]
with \(\lambda \ge 0\) controlling the categorical contribution.

```{r}
iris_mixed <- iris
iris_mixed$Species <- as.character(iris_mixed$Species)
fit_kmm <- cluster(iris_mixed, method = "kmm", k = 3, seed = 1)
fit_kmm
summary(fit_kmm)
```

## Validation metrics

The package reports silhouette width, Calinski-Harabasz, Davies-Bouldin, total
within-cluster sum of squares, bootstrap ARI, ARI, and NMI.

\[
s(i) = \frac{b(i) - a(i)}{\max\{a(i), b(i)\}},
\qquad
\mathrm{CH} = \frac{\mathrm{BSS}/(k - 1)}{\mathrm{WSS}/(n - k)},
\]

\[
\mathrm{DB} = \frac{1}{k} \sum_{j=1}^k \max_{l \ne j} \frac{s_j + s_l}{d(\mu_j,\mu_l)},
\qquad
\mathrm{NMI}(x,y) = \frac{I(x;y)}{\sqrt{H(x)H(y)}}.
\]

```{r}
validate(fit_kmeans)$metrics_table
```

## Exploration and prediction

```{r}
exp <- explore(fit_kmeans)
head(exp$feature_summary)
pred <- predict(fit_kmeans, iris[1:5, 1:4])
pred
```

## Interpretability

`feature_importance()`, `ceteris_paribus()`, and `lime_explain()` all operate
on the fitted clustering rule through `predict()`.

```{r}
feature_importance(fit_kmeans, n_repeats = 3, seed = 1)$summary
```

