---
title: "phynotype: complete workflow"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
  rmarkdown::pdf_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{phynotype: complete workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%"
)
library(phynotype)
set.seed(42)
```

# Introduction

`phynotype` provides tools for unsupervised phenotyping workflows.
The full pipeline has six steps, each with a single entry-point function:

| Step | Function | Purpose |
|------|----------|---------|
| 1 | `cluster()` | Fit a single clustering solution |
| 2 | `metacluster()` | Combine candidates into a consensus |
| 3 | `validate()` | Score a solution with internal/external metrics |
| 4 | `explore()` | Summarize sizes, profiles, and embeddings |
| 5 | `predict()` | Assign new observations |
| 6 | `feature_importance()` / `lime_explain()` / `ceteris_paribus()` | Interpret the clustering rule |

All steps return structured S3 objects with `print()`, `summary()`, and
`plot()` methods. This vignette is the canonical reference and walks through
the complete workflow using the `iris` data set as a running example.

---

# Data preparation

## Numeric data

For purely numeric inputs, pass the matrix or data frame directly to
`cluster()`. Centering and scaling are available via `center` and `scale`:

```{r}
X <- iris[, 1:4]  # 150 x 4 numeric matrix
head(X)
```

## Mixed-type data

When the data contain categorical variables, two strategies are available
depending on the clustering method:

**One-hot encoding** (for k-means, PAM, DBSCAN, GMM):

```{r}
iris_mixed <- iris          # 4 numeric + 1 factor column
iris_mixed$Group <- as.character(
  ifelse(iris$Sepal.Length > 5.8, "large", "small")
)
Xenc <- prepare_mixed_data(iris_mixed[, c(1:4, 6)], center = TRUE, scale = TRUE)
dim(Xenc)
```

**Gower distance** (for hierarchical methods):

```{r}
d_gower <- mixed_distance(iris_mixed[, c(1:4, 6)])
class(d_gower)
```

The Gower dissimilarity between observations $i$ and $j$ is
$$
d_G(i,j) =
\frac{\sum_{f=1}^{p} w_{ijf}\,\delta_{ijf}\,s_{ijf}}
     {\sum_{f=1}^{p} w_{ijf}\,\delta_{ijf}},
$$
where $\delta_{ijf}$ flags comparable pairs, $w_{ijf}$ is an optional
feature weight, and $s_{ijf}$ is the feature-specific partial similarity
(scaled absolute difference for numeric, mismatch indicator for categorical).

---

# Clustering algorithms

## K-means

Lloyd's algorithm (Lloyd, 1982) minimizes the total within-cluster sum of
squares:
$$
\min_{\mathcal{C}} \sum_{j=1}^{k} \sum_{x_i \in C_j} \|x_i - \mu_j\|^2,
\qquad \mu_j = \frac{1}{|C_j|} \sum_{x_i \in C_j} x_i.
$$
The algorithm alternates between nearest-centroid assignment and centroid
recomputation until convergence. Multiple restarts (`nstart`) reduce
sensitivity to random initialization.

```{r}
fit_km <- cluster(X, method = "kmeans", k = 3, seed = 1)
fit_km
```

## PAM

Partitioning Around Medoids (Kaufman and Rousseeuw, 1990) selects $k$ actual
observations (medoids) $m_1, \ldots, m_k$ to minimize
$$
\min_{m_1,\ldots,m_k \in \mathcal{X}}
\sum_{i=1}^{n} \min_{g} d(x_i, m_g).
$$
Unlike k-means, medoids are always observed data points, making PAM more
robust to outliers.

```{r}
if (requireNamespace("cluster", quietly = TRUE)) {
  fit_pam <- cluster(X, method = "pam", k = 3)
  fit_pam
}
```

## Hierarchical clustering

`"hclust"` and `"agnes"` build a dendrogram by successively merging the two
closest clusters. The inter-cluster distance depends on the linkage:

| Linkage | Formula |
|---------|---------|
| Complete | $\max_{a \in A, b \in B} d(a,b)$ |
| Single | $\min_{a \in A, b \in B} d(a,b)$ |
| Average (UPGMA) | $\frac{1}{|A||B|}\sum_{a,b} d(a,b)$ |
| Ward | Minimize the increase in total within-cluster variance |

The final partition is obtained by cutting the tree at height $k$.

```{r}
# From a distance object (supports mixed data via mixed_distance())
d <- mixed_distance(X)
fit_hc <- cluster(d, method = "hclust", k = 3)
fit_hc
plot_dendrogram(fit_hc)
```

AGNES (Kaufman and Rousseeuw, 1990) is the `cluster` package's implementation
and additionally reports the agglomerative coefficient:

```{r}
if (requireNamespace("cluster", quietly = TRUE)) {
  fit_ag <- cluster(d, method = "agnes", k = 3)
  fit_ag
}
```

## DBSCAN

DBSCAN (Ester et al., 1996) groups observations reachable through dense
neighborhoods without requiring $k$. A point $p$ is a *core point* when
$$
|N_\varepsilon(p)| = |\{q : d(p, q) \le \varepsilon\}| \ge \mathrm{MinPts}.
$$
Points not reachable from any core point are labeled *noise* (cluster 0).

```{r}
if (requireNamespace("dbscan", quietly = TRUE)) {
  fit_db <- cluster(X, method = "dbscan", eps = 0.5, minPts = 5)
  fit_db
  # Noise points
  sum(clusters(fit_db) == 0)
}
```

## Gaussian mixture models

A GMM models the data density as a weighted sum of Gaussians:
$$
p(x) = \sum_{g=1}^{k} \pi_g \,\mathcal{N}(x \mid \mu_g, \Sigma_g).
$$
Parameters $(\pi_g, \mu_g, \Sigma_g)$ are estimated by the EM algorithm (Fraley
and Raftery, 2002). Hard labels come from the MAP rule
$\hat{y}_i = \arg\max_g \pi_g \mathcal{N}(x_i \mid \mu_g, \Sigma_g)$; soft
memberships are stored in `membership(fit)`.

```{r}
if (requireNamespace("mclust", quietly = TRUE)) {
  fit_gmm <- cluster(X, method = "gmm", k = 3, seed = 1)
  fit_gmm
  # Soft membership probabilities
  head(membership(fit_gmm))
}
```

## K-prototypes

K-prototypes (Huang, 1998) extends k-means to mixed data by combining squared
Euclidean distance on numeric features with a Hamming mismatch penalty on
categorical features:
$$
D(x_i, p_g) =
\underbrace{\sum_{j \in \mathcal{N}} (x_{ij} - \mu_{gj})^2}_{\text{numeric}}
+\, \lambda
\underbrace{\sum_{j \in \mathcal{C}} I(x_{ij} \neq \nu_{gj})}_{\text{categorical}}.
$$
When `lambda = NULL`, $\lambda$ is estimated automatically from the data.

```{r}
if (requireNamespace("clustMixType", quietly = TRUE)) {
  fit_kp <- cluster(iris, method = "kproto", k = 3, seed = 1)
  fit_kp
  # Mixed prototypes (centroids + modes)
  prototypes(fit_kp)
}
```

## KMM (K-Mixed-Modes)

KMM is `phynotype`'s native mixed-data algorithm. It minimizes the same
weighted prototype distance as k-prototypes:
$$
D(x_i, p_g) = \sum_{j \in \mathcal{N}} (x_{ij} - \mu_{gj})^2
+ \lambda \sum_{j \in \mathcal{C}} I(x_{ij} \neq \nu_{gj}),
$$
with $\lambda \ge 0$ estimated automatically (as the median numeric variance)
when not supplied. KMM uses multiple random restarts and is designed for
biostatistical applications where the number of categories is small relative
to the number of observations.

```{r}
iris_chr <- iris
iris_chr$Species <- as.character(iris_chr$Species)

fit_kmm <- cluster(iris_chr, method = "kmm", k = 3, seed = 1)
fit_kmm
summary(fit_kmm)
prototypes(fit_kmm)
```

---

# Accessors

All `cluster_fit` and `metacluster_fit` objects share a uniform accessor
interface:

```{r}
# Integer assignment vector
head(clusters(fit_km))

# Numeric centroid matrix (k-means, hclust)
centers(fit_km)

# Named size vector
sizes(fit_km)

# Number of clusters
n_clusters(fit_km)

# Method label
method_used(fit_km)
```

---

# Validation

`validate()` computes a table of internal metrics for any `cluster_fit` or
`metacluster_fit`. External metrics are appended when `truth` labels are
supplied.

## Internal metrics

```{r}
val <- validate(fit_km)
val$metrics_table
```

The five internal metrics each capture a different aspect of cluster quality.

**Silhouette width** (Rousseeuw, 1987) measures how similar each observation is
to its own cluster compared to the nearest alternative cluster. For observation
$i$, let $a(i)$ be its mean distance to all other members of its cluster and
$b(i)$ be its mean distance to the closest *other* cluster. The silhouette
width is:
$$
s(i) = \frac{b(i) - a(i)}{\max\{a(i),\, b(i)\}}.
$$
This quantity lies in $[-1, 1]$. A value near $+1$ means $i$ is well inside its
cluster; near $0$ means it sits on a cluster boundary; negative means it is
closer on average to a neighboring cluster than to its own. `validate()` reports
the mean over all observations — a global summary of cluster cohesion and
separation simultaneously.

**Calinski-Harabasz index** (Caliński and Harabasz, 1974) treats clustering as
an ANOVA-like problem: it asks how large the between-cluster dispersion is
relative to the within-cluster dispersion, normalizing by the corresponding
degrees of freedom:
$$
\mathrm{CH} = \frac{\mathrm{BSS} / (k-1)}{\mathrm{WSS} / (n-k)},
$$
where $\mathrm{BSS}$ is the between-cluster sum of squares and $\mathrm{WSS}$
is the total within-cluster sum of squares. The ratio is unbounded above;
higher values indicate tighter, more separated clusters. This index is most
useful when comparing solutions with different values of $k$.

**Davies-Bouldin index** (Davies and Bouldin, 1979) penalizes clusters that are
internally diffuse *and* close to their neighbors. For each cluster $j$, it
finds the worst-case neighbor — the one that is both the most scattered and the
closest — and averages these worst cases:
$$
\mathrm{DB} = \frac{1}{k} \sum_{j=1}^{k} \max_{l \ne j}
\frac{s_j + s_l}{d(\mu_j, \mu_l)},
$$
where $s_j$ is the mean distance from the cluster-$j$ members to centroid
$\mu_j$, and $d(\mu_j, \mu_l)$ is the Euclidean distance between centroids.
Lower values are better; a perfect score of 0 would require clusters with zero
internal scatter.

**Total within-cluster sum of squares** is the raw objective minimized by
k-means, and is included as a direct measure of compactness:
$$
\mathrm{WSS} = \sum_{j=1}^{k} \sum_{i \in C_j} \|x_i - \mu_j\|^2.
$$
It is most interpretable when compared across solutions with the same $k$
— the "elbow" in a WSS-vs-$k$ plot signals a natural cluster count.

**Bootstrap ARI** (Fang and Wang, 2012) addresses stability rather than
geometric quality. It refits the same clustering model on $B$ bootstrap
resamples of the data and asks: how similar are the resulting partitions to the
original? Formally,
$$
\mathrm{BootARI} = \frac{1}{B} \sum_{b=1}^{B}
\mathrm{ARI}\!\left(\hat{y}_{I^{(b)}},\, \hat{y}^{(b)}\right),
$$
where $I^{(b)}$ is the set of indices drawn in bootstrap $b$, $\hat{y}^{(b)}$
is the partition refitted on those indices, and ARI is evaluated only on the
sampled observations. Values near 1 mean the cluster structure is robust to
resampling; values below 0.6 suggest the solution is sensitive to the
particular data points included.

## External metrics

When reference labels are available:

```{r}
val_ext <- validate(fit_km, truth = iris$Species)
val_ext$metrics_table
```

**Adjusted Rand index** (Hubert and Arabie, 1985) counts the proportion of
observation pairs on which two labelings agree — both in the same cluster, or
both in different clusters — and corrects that count for the level of agreement
expected by chance alone. Given a contingency table $(n_{ij})$ with row sums
$a_i$ and column sums $b_j$:
$$
\mathrm{ARI}(U, V) = \frac{
\sum_{ij}\binom{n_{ij}}{2} - E
}{
\tfrac{1}{2}\!\left[\sum_i\binom{a_i}{2}+\sum_j\binom{b_j}{2}\right] - E
},
\quad E = \frac{\sum_i\binom{a_i}{2}\cdot\sum_j\binom{b_j}{2}}{\binom{n}{2}}.
$$
A value of 1 indicates perfect agreement; the expected value under random
labelings is 0; negative values are possible when agreement is worse than
chance.

**Normalized mutual information** (Strehl and Ghosh, 2002) measures how much
knowing one partition reduces uncertainty about the other, normalized so that
the scale is comparable across different cluster counts:
$$
\mathrm{NMI}(U, V) = \frac{I(U; V)}{\sqrt{H(U)\,H(V)}},
$$
where $I(U; V) = \sum_{u,v} p_{uv} \log(p_{uv} / p_u p_v)$ is mutual
information and $H(U) = -\sum_u p_u \log p_u$ is Shannon entropy. Unlike ARI,
NMI is not corrected for chance, but it is symmetric and bounded in $[0, 1]$,
making it useful for reporting alongside ARI.

## Grid search

To compare cluster counts, pass raw data directly:

```{r}
grid_val <- validate(X, method = "kmeans", k = 2:6)
grid_val$metrics_table
```

## Per-cluster silhouette widths

Individual silhouette widths reveal which observations are well-assigned
(positive width, comfortably inside their cluster) and which are borderline
or potentially misclassified (width near zero or negative):

```{r}
if (requireNamespace("cluster", quietly = TRUE)) {
  plot_silhouette(fit_km)
}
```

---

# Exploration

`explore()` computes four structural summaries from a fitted clustering object.

```{r}
exp <- explore(fit_km)
```

**Cluster sizes**:

```{r}
exp$size_table
plot_cluster_sizes(fit_km)
```

**Feature profiles** (per-cluster mean, SD, median, min, max):

```{r}
head(exp$feature_summary)
plot_feature_profiles(exp)
```

**Feature separation** (eta-squared):
$$
\eta^2_j = \frac{\mathrm{SS}_{B,j}}{\mathrm{SS}_{T,j}}
= \frac{\sum_{g=1}^{k} n_g (\bar{x}_{gj} - \bar{x}_j)^2}
       {\sum_{i=1}^{n} (x_{ij} - \bar{x}_j)^2}.
$$
Values near 1 indicate features that strongly discriminate clusters.

```{r}
exp$separation_table
```

**PCA embedding**:

```{r}
plot_clusters(fit_km)
```

---

# Prediction

`predict()` assigns new observations to the nearest cluster center or
prototype. The prediction rule is method-dependent: nearest centroid for
k-means and hierarchical fits, nearest medoid for PAM, density-based for
DBSCAN, and MAP for GMM.

```{r}
new_obs <- data.frame(
  Sepal.Length = c(5.0, 6.5, 7.2),
  Sepal.Width  = c(3.5, 2.9, 3.2),
  Petal.Length = c(1.5, 4.5, 6.0),
  Petal.Width  = c(0.3, 1.5, 2.2)
)

pred <- predict(fit_km, new_obs)
pred$clusters
```

Distance to each cluster center:

```{r}
head(pred$distances)
```

---

# Consensus meta-clustering

When no single algorithm or value of $k$ is clearly best, `metacluster()`
pools candidate solutions into a consensus through co-association evidence
accumulation (Fred and Jain, 2002).

## How it works

For $B$ candidate partitions, the co-association matrix $C \in [0,1]^{n\times n}$
is
$$
C_{ij} = \frac{1}{B} \sum_{b=1}^{B}
I\!\left\{u^{(b)}_i = u^{(b)}_j\right\},
$$
the proportion of candidates that co-cluster observations $i$ and $j$. The
consensus dissimilarity $D = 1 - C$ is then hierarchically clustered (average
linkage) and the optimal $k$ is chosen by maximizing the mean silhouette width:
$$
k^* = \operatorname*{arg\,max}_{k \in \mathcal{K}} \bar{s}_k(D).
$$

Ensemble stability is reported as the mean pairwise partition agreement (PPA):
$$
\mathrm{PPA}(U, V) = \frac{1}{\binom{n}{2}}
\sum_{i < j} I\!\left\{(u_i = u_j) = (v_i = v_j)\right\}.
$$

## Usage

```{r}
mfit <- metacluster(
  X,
  methods = c("kmeans", "pam", "hclust"),
  k       = 2:5,
  seed    = 1
)
mfit
```

Candidate partition table:

```{r}
head(mfit$candidate_table)
```

Ensemble stability:

```{r}
mfit$stability_summary
```

Consensus $k$ selection scores:

```{r}
mfit$selection_summary
```

## Validation and exploration of the consensus

```{r}
validate(mfit)$metrics_table
plot_coassoc(mfit)
plot_consensus(mfit)
```

Consensus dendrogram:

```{r}
plot_dendrogram(mfit)
```

---

# Interpretation

## The unsupervised interpretation problem

Classical interpretability methods — permutation importance, LIME, individual
conditional profiles — were designed for supervised models where a response
variable $y$ is available. Clustering produces no such variable. The challenge
is that we want to understand the clustering *rule* $f: \mathcal{X} \to C$,
not the cluster structure in the data.

`phynotype` resolves this by treating the fitted model as a black-box
predictor. Every supported method with a `predict()` implementation maps new
observations to cluster labels (and optionally to distance or membership
scores). This map is well-defined, deterministic, and can be probed exactly
like any classifier. The three interpretability functions below operate
exclusively through this map — they never look at the raw training data or the
clustering criterion directly. As a result, the explanations describe **the
decision boundary of the fitted model**, not the geometry of the original
point cloud. This distinction matters: a feature can be highly informative for
separating the data yet contribute little to the fitted boundary if the
algorithm did not use it, and vice versa.

All three functions require a fit trained on row-by-feature data with
prediction support (`"kmeans"`, `"pam"`, `"hclust"`, `"agnes"`, `"gmm"`,
`"kproto"`, `"kmm"`).

## Global feature importance

**What it answers:** which features does the fitted clustering rule rely on
most, globally across all observations?

The approach adapts permutation importance (Breiman, 2001; Fisher et al.,
2019) to the unsupervised setting. The idea is simple: if a feature is
important to the model, randomly shuffling its values across observations
should disrupt the cluster assignments. If it is irrelevant, shuffling should
leave assignments unchanged.

Concretely, let $\hat{c} = f(X)$ be the baseline partition produced by the
fitted model on the evaluation data $X$. For each feature $j$, we independently
permute its column $M$ times and re-predict:
$$
\mathrm{FI}_j = \frac{1}{M} \sum_{m=1}^{M}
\frac{1}{n} \sum_{i=1}^{n}
I\!\left\{\hat{c}_i \neq f\!\left(X^{(m)}_{\pi(j)}\right)_i\right\}.
$$
Here $X^{(m)}_{\pi(j)}$ is the data matrix with column $j$ replaced by a
random permutation of its values. The importance $\mathrm{FI}_j$ is the mean
fraction of observations whose predicted cluster changes when column $j$ is
scrambled. A value of 0 means the model ignores feature $j$ entirely; a value
of 1 means every assignment changes when $j$ is removed.

Two alternative metrics measure the same idea through internal scores rather
than label changes. With `metric = "silhouette"`, importance is the mean drop
in silhouette width after permutation:
$$
\mathrm{FI}_j^{\mathrm{sil}} = \frac{1}{M} \sum_{m=1}^{M}
\left[S\!\left(f, X\right) - S\!\left(f, X^{(m)}_{\pi(j)}\right)\right],
$$
so a positive value means permuting $j$ degrades within-cluster cohesion.
With `metric = "total_within"`, it is the mean increase in within-cluster
dispersion, capturing the same idea from a compactness perspective.

```{r}
imp <- feature_importance(fit_km, n_repeats = 5, seed = 1)
imp$summary
plot(imp)
```

The bar chart ranks features by their mean importance. In this k-means fit on
`iris`, `Petal.Length` and `Petal.Width` dominate — permuting either of them
strongly disrupts which cluster each observation is sent to, confirming that
the fitted boundary is primarily organised along the petal dimensions.

## Local explanations (LIME)

**What it answers:** for a specific observation, which features pushed it into
its assigned cluster, and in which direction?

Global importance tells us which features matter on average across all
observations, but it cannot explain an individual assignment. LIME (Local
Interpretable Model-agnostic Explanations; Ribeiro et al., 2016) addresses
this by fitting a simple linear model locally around each observation being
explained.

In the supervised setting LIME explains a classifier's predicted class. Here
the "classifier" is the clustering rule $f$. For each explained observation
$x$, we define a target $g\{f(\cdot)\}$ that converts model output into a
scalar:

- With `target = "cluster"` (default): $g\{f(z)\} = I\{f(z) = c\}$, a binary
  indicator of whether the perturbed sample $z$ is predicted to belong to the
  same cluster $c$ as the original observation. The local linear model then
  identifies which features explain membership in that cluster.
- With `target = "score"`: $g\{f(z)\}$ is the cluster-specific membership
  probability (for GMM) or a normalized radial similarity score derived from
  distances (for centroid/prototype methods). This gives a continuous response
  and is appropriate when soft assignment information is meaningful.

The LIME objective is to find a sparse linear surrogate $\hat{h}$ that
approximates $g\{f(\cdot)\}$ in a weighted neighborhood around $x$:
$$
\hat{h} = \operatorname*{arg\,min}_{h \in \mathcal{H}}
\sum_{i=1}^{N} \pi_x(z_i)
\left[g\!\left\{f(z_i)\right\} - h(z_i)\right]^2 + \Omega(h),
$$
where $z_1, \ldots, z_N$ are perturbed samples drawn around $x$ (numeric
features are perturbed by Gaussian noise calibrated to their training
standard deviation; categorical features are randomly resampled from observed
levels), and $\pi_x(z_i) = \exp(-\|x - z_i\|^2 / \sigma^2)$ is an
exponential kernel that down-weights samples far from $x$. Sparsity
$\Omega(h)$ is enforced operationally by returning only the top `n_features`
absolute effects from the fitted weighted linear model.

The output is a signed coefficient for each feature: a positive coefficient
means that increasing the feature value makes membership in cluster $c$ more
likely under the local linear approximation; a negative coefficient means the
opposite. These are *local* effects — they describe the boundary in the
immediate neighborhood of the explained observation and should not be
extrapolated globally.

```{r}
lx <- lime_explain(fit_km, iris[c(1, 51, 101), 1:4],
                   n_permutations = 100, n_features = 4, seed = 1)
lx$explanations[, c("observation", "feature", "estimate", "direction")]
plot(lx)
```

The plot shows one panel per explained observation. Reading the bars:
observation 1 (setosa, cluster 1) is explained primarily by small
`Petal.Length` pushing it away from the other clusters; observation 51
(versicolor) and observation 101 (virginica) are separated mainly along
`Petal.Width` and `Petal.Length`, consistent with the global importance
result.

## Ceteris paribus profiles

**What it answers:** how does the predicted cluster assignment (or cluster
score) change as one feature varies, holding everything else fixed at its
observed value for a specific observation?

LIME gives a linear summary of the local boundary. Ceteris paribus (CP)
profiles give the full nonlinear picture for each feature of interest. The
name comes from the Latin for "all other things being equal": for observation
$x_i$ and feature $j$, we slide $j$ across a grid of values
$z_1, \ldots, z_G$ while clamping all other features at their observed values,
then record the model's prediction at each grid point:
$$
\mathrm{CP}_{i,j}(z) = g\!\left\{f\!\left(x_{i,-j},\, z\right)\right\},
$$
where $x_{i,-j}$ denotes the feature vector of observation $i$ with feature
$j$ replaced by $z$, $f$ is the fitted clustering rule, and $g$ extracts
either the cluster label (producing a step function, `target = "cluster"`) or
a continuous cluster-specific score (`target = "score"`).

CP profiles answer the question: "if this particular patient had a different
value for `Petal.Length`, would their phenotype assignment change, and at what
threshold?" They are individual diagnostic curves — useful for understanding
borderline cases and for communicating individual-level model behaviour to
domain experts.

```{r}
cp <- ceteris_paribus(
  fit_km,
  iris[c(1, 51, 101), 1:4],
  features  = c("Petal.Length", "Petal.Width"),
  grid_size = 20
)
plot(cp)
```

Each panel shows one feature; each line corresponds to one explained
observation. The dashed vertical line marks the observation's actual feature
value, and the dot marks its baseline prediction. Step changes in the profile
(for `target = "cluster"`) reveal the exact feature thresholds at which the
model switches cluster assignment — directly interpretable as phenotype
decision boundaries.

---

# Complete pipeline example

The full pipeline in compact form:

```{r, eval=FALSE}
# 1. Fit
fit <- cluster(iris[, 1:4], method = "kmeans", k = 3, seed = 1)

# 2. Validate
val <- validate(fit, truth = iris$Species)

# 3. Explore
exp <- explore(fit)

# 4. Predict
pred <- predict(fit, iris[1:5, 1:4])

# 5. Interpret
imp <- feature_importance(fit, n_repeats = 10, seed = 1)
lx  <- lime_explain(fit, iris[1:3, 1:4], n_permutations = 200, seed = 1)
cp  <- ceteris_paribus(fit, iris[1:3, 1:4])

# 6. Consensus (when a single solution is uncertain)
mfit <- metacluster(iris[, 1:4],
                    methods = c("kmeans", "pam", "hclust"),
                    k = 2:5, seed = 1)
validate(mfit)$metrics_table
```

---

# References

Apley, D.W. and Zhu, J. (2020). Visualizing the effects of predictor variables
in black box supervised learning models. *Journal of the Royal Statistical
Society: Series B*, **82**(4), 1059--1086.

Biecek, P. and Burzykowski, T. (2021). *Explanatory Model Analysis*.
Chapman and Hall/CRC, Boca Raton. <https://ema.drwhy.ai/>

Breiman, L. (2001). Random forests. *Machine Learning*, **45**(1), 5--32.

Caliński, T. and Harabasz, J. (1974). A dendrite method for cluster analysis.
*Communications in Statistics*, **3**(1), 1--27.

Davies, D.L. and Bouldin, D.W. (1979). A cluster separation measure. *IEEE
Transactions on Pattern Analysis and Machine Intelligence*, **1**(2), 224--227.

Ester, M., Kriegel, H.-P., Sander, J. and Xu, X. (1996). A density-based
algorithm for discovering clusters in large spatial databases with noise.
*Proceedings of the 2nd ACM SIGKDD*, pp. 226--231.

Fang, Y. and Wang, J. (2012). Selection of the number of clusters via the
bootstrap method. *Computational Statistics and Data Analysis*, **56**(3),
468--477.

Fisher, A., Rudin, C. and Dominici, F. (2019). All models are wrong, but many
are useful: Learning a variable's importance by studying an entire class of
prediction models simultaneously. *Journal of Machine Learning Research*,
**20**(177), 1--81.

Fraley, C. and Raftery, A.E. (2002). Model-based clustering, discriminant
analysis, and density estimation. *Journal of the American Statistical
Association*, **97**(458), 611--631.

Fred, A.L.N. and Jain, A.K. (2002). Data clustering using evidence
accumulation. *Proceedings of the 16th International Conference on Pattern
Recognition (ICPR'02)*, Vol. 4, pp. 276--280.

Gosiewska, A. and Biecek, P. (2019). iBreakDown: Uncertainty of model
explanations for non-additive predictive models. *arXiv:1903.11420*.

Gower, J.C. (1971). A general coefficient of similarity and some of its
properties. *Biometrics*, **27**(4), 857--874.

Huang, Z. (1998). Extensions to the k-means algorithm for clustering large
data sets with categorical values. *Data Mining and Knowledge Discovery*,
**2**(3), 283--304.

Hubert, L. and Arabie, P. (1985). Comparing partitions. *Journal of
Classification*, **2**(1), 193--218.

Kaufman, L. and Rousseeuw, P.J. (1990). *Finding Groups in Data: An
Introduction to Cluster Analysis*. John Wiley & Sons, New York.

Lloyd, S.P. (1982). Least squares quantization in PCM. *IEEE Transactions on
Information Theory*, **28**(2), 129--137.

MacQueen, J. (1967). Some methods for classification and analysis of
multivariate observations. *Proceedings of the 5th Berkeley Symposium on
Mathematical Statistics and Probability*, Vol. 1, pp. 281--297.

Murtagh, F. and Legendre, P. (2014). Ward's hierarchical agglomerative
clustering method: Which algorithms implement Ward's criterion? *Journal of
Classification*, **31**(3), 274--295.

Ribeiro, M.T., Singh, S. and Guestrin, C. (2016). "Why should I trust you?":
Explaining the predictions of any classifier. *Proceedings of the 22nd ACM
SIGKDD International Conference on Knowledge Discovery and Data Mining*,
pp. 1135--1144.

Rousseeuw, P.J. (1987). Silhouettes: A graphical aid to the interpretation and
validation of cluster analysis. *Journal of Computational and Applied
Mathematics*, **20**, 53--65.

Strehl, A. and Ghosh, J. (2002). Cluster ensembles: A knowledge reuse
framework for combining multiple partitions. *Journal of Machine Learning
Research*, **3**, 583--617.

Ward, J.H. (1963). Hierarchical grouping to optimize an objective function.
*Journal of the American Statistical Association*, **58**(301), 236--244.
